---
title: "Lung Cancer Survival Dashboard"
author: "Neil Gutlay"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
---

```{r setup, include=FALSE}
library(flexdashboard)
library(survival)
library(survminer)
library(ggplot2)
library(dplyr)
library(DT)
library(plotly)
library(broom)
```

```{r data, include=FALSE}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("VectraPolarisData")

library(VectraPolarisData)
spe_lung <- HumanLungCancerV3()

assays_slot <- assays(spe_lung)
intensities_df <- assays_slot$intensities
rownames(intensities_df) <- paste0("total_", rownames(intensities_df))
nucleus_intensities_df <- assays_slot$nucleus_intensities
rownames(nucleus_intensities_df) <- paste0("nucleus_", rownames(nucleus_intensities_df))
membrane_intensities_df <- assays_slot$membrane_intensities
rownames(membrane_intensities_df) <- paste0("membrane_", rownames(membrane_intensities_df))

colData_df <- colData(spe_lung)
spatialCoords_df <- spatialCoords(spe_lung)
patient_level_df <- metadata(spe_lung)$clinical_data

cell_level_df <- as.data.frame(cbind(colData_df, 
                                     spatialCoords_df,
                                     t(intensities_df),
                                     t(nucleus_intensities_df),
                                     t(membrane_intensities_df)))

lung_df <- full_join(patient_level_df, cell_level_df, by = "slide_id")

lung_df_2 <- lung_df %>%
  select(slide_id, gender, mhcII_status, age_at_diagnosis,
         stage_at_diagnosis, stage_numeric, pack_years, survival_days,
         survival_status, cause_of_death, adjuvant_therapy,
         time_to_recurrence_days, recurrence_or_lung_ca_death)

data <- lung_df_2 %>% distinct()
```

Survival Analysis {.tabset}
===========================================================

Column {data-width=650}
--------------------------------

### Kaplan-Meier Survival Curves

```{r km_plot}
library(survival)
library(broom)
library(ggplot2)
library(plotly)
library(dplyr)

# Step 1: Filter out missing data
km_data <- data %>%
  filter(!is.na(survival_days),
         !is.na(survival_status),
         !is.na(mhcII_status))

# Step 2: Fit survival model
fit <- survfit(Surv(survival_days, survival_status) ~ mhcII_status, data = km_data)

# Step 3: Tidy and clean up strata names
tidy_km <- broom::tidy(fit) %>%
  mutate(strata = gsub("mhcII_status=", "", strata))

# Step 4: Plot survival curves
km_plot <- ggplot(tidy_km, aes(x = time, y = estimate, color = strata,
                               group = strata,
                               text = paste0("Time: ", time,
                                             "<br>Survival Prob: ", round(estimate, 3),
                                             "<br>Events: ", n.event,
                                             "<br>At Risk: ", n.risk))) +
  geom_step(size = 1.2) +
  labs(title = "Kaplan-Meier Survival Curves by MHC-II Status",
       x = "Time (Days)",
       y = "Survival Probability",
       color = "MHC-II Status") +
  theme_minimal()

# Step 5: Convert to interactive plot
ggplotly(km_plot, tooltip = "text")
```

> The Kaplan-Meier survival curve estimates the probability of survival over time for different patient groups.
Here, we compare survival probabilities based on MHC-II status.
The risk table shows the number of patients remaining at risk at different time points.
The p-value from the log-rank test suggests there is no statistically significant difference between groups.

Column
--------------------------------

### Cox Proportional Hazards Model

```{r cox_model}
# Step 1: Fit Cox model
cox_fit <- coxph(Surv(survival_days, survival_status) ~ 
                   pack_years + mhcII_status + stage_at_diagnosis, data = data)

# Step 2: Tidy the results for plotting
tidy_cox <- tidy(cox_fit, exponentiate = TRUE, conf.int = TRUE) %>%
  mutate(term = gsub("_", " ", term))  # Optional: clean variable names

# Step 3: Build interactive forest plot
forest_plot <- ggplot(tidy_cox, aes(x = estimate, y = reorder(term, estimate),
                                    xmin = conf.low, xmax = conf.high,
                                    text = paste0("Variable: ", term,
                                                  "<br>HR: ", round(estimate, 2),
                                                  "<br>95% CI: [", round(conf.low, 2), ", ", round(conf.high, 2), "]",
                                                  "<br>p-value: ", signif(p.value, 3)))) +
  geom_point(size = 3, color = "#2c7fb8") +
  geom_errorbarh(height = 0.2, color = "#2c7fb8") +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  labs(title = "Cox Proportional Hazards Model",
       x = "Hazard Ratio (HR)",
       y = NULL) +
  theme_minimal()

# Step 4: Make it interactive
ggplotly(forest_plot, tooltip = "text")

```

> The Cox proportional hazards model estimates the effect of various clinical and biological factors on survival risk. Each hazard ratio (HR) represents the relative risk of death associated with a given variable. Values greater than 1 indicate an increased risk, while values less than 1 suggest a protective effect. Confidence intervals help assess the reliability of each estimate.

Clinical Dataset {.tabset}
===========================================================

Column
--------------------------------

### Interactive DataTable

```{r data_table}
datatable(data, filter = 'top', options = list(pageLength = 10))
```
